Building socket-aware BPF programs

### 2018-10-27 paper/paper.tex
paper: Flesh out general structure and some text

    This paper presents the changes necessary to extend socket introspection
    capabilities of the Linux kernel's BPF packet hooks. To provide safety
    guarantees around the access of the socket, the verifier is extended to
    track allocation and release of kernel resources. Using this mechanism, we
    introduce socket lookup and release helpers into the BPF API.

    %These new primitives open up new possibilities for various use cases
    %including stateful firewalling and load balancing.

@@

\begin{itemize}
\item BPF started with sockets, but only provided access to packets
\item Previous helper extensions allowed e.g. socket operation handling
\item Arbitrary socket lookup from socket/packet lookups allows more powerful
      BPF programs
\end{itemize}

\todo{Rewrite introduction}

@@

%To build such awareness into a BPF helper, the verifier needs the ability to
%track the safety of the calls, including appropriate access into socket
%pointer types, and reference counting upon the underlying socket. This paper
%walks through extensions to the verifier to perform tracking of references in
%a BPF program. This allows BPF developers to extend the API with functions
%that allocate and release resources within the execution lifetime of a BPF
%program, and the verifier will validate that the resource acquisition and
%release is balanced across all execution paths of the BPF programs.

%Using this new reference tracking ability in the verifier, we add socket lookup
%and release function calls to the BPF API, allowing BPF programs to safely find
%a socket and build logic upon the presence or attributes of a socket. This can
%be used to load-balance traffic based on the presence of a listening
%application, or to implement stateful firewalling primitives to understand
%whether traffic for this connection has been seen before. With this new
%functionality, BPF programs can integrate more closely with the networking
%stack's understanding of the traffic transiting the kernel.

\todo{Desire to target logic based upon socket existence/attributes at packet
layers}

@@

%To validate access around sockets, the verifier must have knowledge of the
%socket type to validate that access is within bounds and is safe with respect
%to read or write.

The BPF verifier must ensure that all BPF programs that are loaded into the
kernel are safe to execute. For accessing socket pointers, this requires
validation that memory accesses are within the bounds of the memory allocated
for the structure, and will not make any modifications to the socket fields
that would be incompatible with the core socket handling logic. This is
provided through extension of existing functionality which provides bounds
verification and offset conversion for programs operating on CGroups hooks.
Additionally, the underlying socket memory must remain associated with the
socket for the duration of accesses from the BPF program, which may be
guaranteed by taking a reference on sockets for the duration of their use.

@@

Earlier work~\cite{bpf-sock} introduced a \verb+bpf_sock+ structure to the BPF
API and implemented bounds and access type checking, along with offset
rewriting for converting the access of socket attributes from BPF instructions
into equivalent accesses of the underlying kernel types. This work was
introduced with new BPF hook points which provide the socket structure as the
context of the BPF program.

To allow BPF programs to retrieve and access a socket pointer, the verifier
must be made aware of when a register contains a pointer of this type, and also
how to validate pointer access. The existing verifier logic handles context
pointers in a generic manner via \verb+bpf_verifier_ops+. However, the packet
event hook points targeted in this paper already contain a context pointer of a
type that is different from \verb+bpf_sock+, so some additional logic was
required in the verifier to understand this pointer type. The patch series
associated with this paper introduced a new pointer type specific to sockets,
and linked its validation directly to the aforementioned socket verification
functions without using the verifier operations abstraction. Write operations
into the socket are rejected in the current implementation.

With the verifier now aware of socket pointer types, this could be built upon
to implicitly associate socket pointers with references to kernel resources.

@@

As with many other structure types, Linux sockets are reference-counted to
track usage and handle memory release. In general to ensure the safety of
socket for the duration of access in BPF programs, this reference count needs
to be incremented when sockets are referenced from BPF programs. Two options
were considered for how to guarantee this: Implicit reference tracking and
explicit reference tracking.

\subsubsection{Implicit reference tracking.}

This calls for the BPF infrastructure to handle reference tracking, and to hide
this detail from the BPF API. Whenever a socket lookup function is called, a
reference is taken and the socket is added to a reference list. At the end of
the BPF program execution, the core code could walk this reference list and
release each reference. However, collecting references and releasing them at
the end of the BPF program invocation has the unfortunate overhead that even
the execution of BPF programs that do not make use of the new socket lookup
helpers would need to pay the cost of checking the list of references to free.
When dealing with sufficiently demanding use cases such as those which use XDP,
even the few instructions required to implement such a check may have a
measurable impact on performance.

\subsubsection{Explicit reference tracking.}

This calls for reference acquisition and release semantics to be built into the
API. If references must be taken, then requiring BPF program writers to
explicitly handle these references ensures that the program writer must
understand the potential impact this may have on the operation of the program
(including atomic operation cost), and it also ensures that the cost is only
borne by programs that make use of this feature.

\subsubsection{Implementation.}

Based upon the tradeoffs described above, explicit reference tracking was
chosen. From an implementation perspective, this requires tracking pointer
values through each conditional path in the program, and rejecting the load of
programs that fail to "balance" resource acquisition (lookup) with release.
This implementation works as follows.

Calls to helper functions which acquire references to resources are annotated
in the verifier to associate the acquisition and release functions. When the
instruction that acquires a resource is processed, a resource identifier is
allocated for this resource. This identifier is kept in a list in the verifier
state, and it is also associated with the register which receives the resulting
resource pointer. When processing a corresponding release helper call, if a
parameter to the function contains a register that is associated with a
resource identifier, then the identifier is removed from the verifier state. If
any paths in the program reach the final (\verb+BPF_EXIT+) instruction and the
verifier state contains any resource identifier association, then the program
is considered unsafe as it has leaked the resource.

Some instruction calls are restricted while holding a resource reference to
avoid leaking references, for example the \verb+bpf_tail_call+ helper function.

@@

The verifier is tasked with ensuring that all BPF program execution flow is
safe to the extents where it interacts with the BPF API. In general, BPF
programs, once verified and triggered by event hooks, will execute from the
first instruction through the final instruction for a particular path through
the program. Therefore, after execution of a BPF program containing resource
acquire and release functions which are verified by the logic above, the
underlying resources should not leak.

One exception to this is when classic BPF \verb+LD_ABS+ or \verb+LD_IND+
instructions are used in the BPF program. These instructions were previously
used to provide direct packet access, with the semantics that if the access
offset exceeds the length of the packet, the BPF program would be terminated
prior to the final state. Without mitigation, this could lead to leaking of
references. Newer BPF APIs provide better alternatives for packet access, so we
disallow the use of these instructions while holding a socket reference.

@@

To implement explicit reference tracking, the API required at least two
functions to handle referencing of sockets from the BPF API: A socket lookup
function and a socket release function. This section describes the API
considerations for each of these.

\subsection{API definition}

The following considerations were made for implementing the lookup helpers. The
consideration of each of these items reflects the definition of the lookup
helpers in Listing~\ref{lst:helper-api}.

\begin{itemize}
    \item The Linux stack may contain multiple network namespaces, where the
          BPF program may operate attached to a device in one network namespace
          while the desired socket may exist in another network namespace.
    \item BPF programmers may wish to discover sockets that are not directly
          related to the BPF program context (packet, socket, etc).
    \item Current use cases are focused around UDP and TCP sockets, however
          there may be a desire to extend this in future.
\end{itemize}

\todo{Span this listing across two columns}
\begin{lstlisting}[caption={BPF API helper functions for socket lookup},language=c,label=lst:helper-api,breaklines=true]
struct bpf_sock *bpf_sk_lookup_tcp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
                                   u32 netns, u64 flags)
struct bpf_sock *bpf_sk_lookup_udp(void *ctx, struct bpf_sock_tuple *tuple, u32 tuple_size,
                                   u32 netns, u64 flags)
int bpf_sk_release(struct bpf_sock *sk)
\end{lstlisting}

\subsubsection{Namespacing.}

Linux network devices are associated with a particular network namespace which
provides logical network stack separation. Sockets are also inherently
associated with an application that operates in a particular network namespace.
The simplest path for handling namespacing from the lookup handlers would be to
allow BPF programs operating on a device within a network namespace to only
find sockets in the same network namespace. By passing the BPF program context
into the socket lookup helper, the implementation can derive the source network
namespace from this context. Beyond this, with the growing popularity of
containers it may be desirable for a network orchestration tool to make use of
socket properties inside container network namespaces to influence decisions
made on packets outside the container---for instance, on packets being received
into a node prior to the packet passing into the container network namespace,
or if filtering is implemented on a veth pair outside of a container. By
allowing the netns ID to be specified, this use case can be supported as well.

\subsubsection{Tuple definition.}

A common use case for the socket lookup functions would call them from a packet
context, where the \verb+__sk_buff+ has the full 5-tuple information available
and easily accessible. One could imagine a simpler helper function that allows
the BPF program writer to provide the \verb+__sk_buff+ to the helper function,
then the helper function would look up the socket based upon the packet
metadata. This would however limit the potential uses for such a helper. Two
cases that would be more limited with this model are usage from XDP, or if the
BPF program implements a form of network address translation. From XDP, if the
raw \verb+xdp_md+ context is provided to such a helper, then the helper would
need to parse the packet to understand the 5-tuple. This would duplicate the
standard stack paths which are executed after XDP where the \verb+sk_buff+ is
built, but this information would be subsequently thrown away upon return of
the helper call. Adding to this, if the program performs network address
translation between the local application and the remote destination, then the
local stack may not contain a socket associated with the tuple that is directly
in the packet. While these cases could be worked around with a simpler helper,
it was deemed more powerful and generic to allow the BPF program writer to
provide the tuple for lookup.

\subsubsection{Extensibility.}

New kernel APIs that have any scope for extension should contain a flags
argument. For the socket lookup helpers, a few ideas had been considered as
potential future alternatives to the existing behaviour. The lookup functions
follow the standard socket lookup paths in Linux which have predetermined
methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
Some discussion on the mailinglist proposed allowing BPF program writers to
influence the socket selection mechanism~\cite{netdev-sk-select}.

\subsubsection{Result.}

Another aspect of extensibility is the ability to lookup sockets which are not
TCP or UDP. The initial RFC of this patch series proposed a single lookup
function which would choose the L4 protocol based upon a field in the
tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
program writers to detect the support for different protocol types at compile
time. There is not a significant number of supported L4 socket types in the
Linux stack today, and the number is not expected to increase drastically, so
it was considered simpler and easier to split out each L4 socket lookup
function into an independent helper call.

\subsection{Optimizations}

Multiple optimizations were proposed during development of this feature to
reduce the runtime overhead of using the socket lookup helpers. Two such
optimizations are described below: Skipping reference counting when unnecessary
to save on atomic instructions, and allowing the use of direct packet pointers
for the tuple.

\subsubsection{Avoid reference counting when unnecessary.}

For some socket types, such as UDP sockets, or TCP listen sockets, memory
access safety can be achieved with minimal implementation: The destruction of
such sockets is already governed by standard RCU rules, meaning that while the
RCU lock is held, they can be safely referenced; Once the RCU grace period is
reached, the memory may be freed and references to the socket are no longer
safe to use. Because BPF programs run under the RCU lock, this allows simple,
safe access of properties of these sockets without requiring additional
mechanisms to ensure the safety of the socket pointers.

When it comes to TCP sockets that are not governed by RCU, the networking stack
uses reference counting to manage socket memory instead. To ensure that the
socket remains safe to access for a period of BPF program execution, the BPF
program must also take a reference on the socket. This also implies that such
references must be subsequently released to prevent leaking kernel resources.

\subsubsection{Allow lookup using direct packet pointers.}

The tuple structure is defined in such a way that if the IPv4 or IPv6 packet is
immediately followed by the TCP or UDP header without IP options in between,
then a pointer directly to the L3 addresses may be passed to the lookup
function, allowing the helper function to directly pull the addresses and ports
from the packet buffer rather than requiring the BPF program to first extract
these onto the stack.

\subsection{Future work}

The implementation introduced in the 4.20 release cycle includes support for
looking up TCP and UDP sockets for IPv4 and IPv6 traffic, using the
\verb+SCHED_CLS+, \verb+SCHED_ACT+ and \verb+SK_SKB+ hook points. A proposed
patchset extends this to allow the same helpers to be used from the \verb+XDP+
hook points~\cite{sk-lookup-xdp}.

\todo{Associate custom metadata with socket?}

@@

This paper describes the contribution of new BPF verifier functionality in the
Linux kernel to track references to kernel resources and uses these to provide
access to socket introspection capabilities from packet handling hooks. The
introduction of reference tracking logic into the verifier provides
useful base infrastructure for supporting acquire and release semantics for
kernel resources which may prove useful for other helpers in future, and the
socket lookup API makes the packet hooks more powerful to support use cases
such as stateful firewalling or load-balancing based on the sockets that are
open on a system.

@@

The core BPF maintainers provided great guidance and review of the patches, in
particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.

>>>>

    This paper presents the changes necessary to extend socket introspection
    capabilities of the Linux kernel's BPF packet hooks. To provide safety
    guarantees around the access of the socket, the verifier was extended to
    track allocation and release of kernel resources. Using this mechanism, we
    introduce socket lookup and release helpers into the BPF API.

@@

@@

To build such awareness into a BPF helper, the verifier needs the ability to
track the safety of the calls, including appropriate reference counting upon
the underlying socket. This talk walks through extensions to the verifier to
perform tracking of references in a BPF program. This allows BPF developers to
extend the UAPI with functions that allocate and release resources within the
execution lifetime of a BPF program, and the verifier will validate that the
resources are released exactly once prior to program completion.

Using this new reference tracking ability in the verifier, we add socket lookup
and release function calls to the BPF API, allowing BPF programs to safely find
a socket and build logic upon the presence or attributes of a socket. This can
be used to load-balance traffic based on the presence of a listening
application, or to implement stateful firewalling primitives to understand
whether traffic for this connection has been seen before. With this new
functionality, BPF programs can integrate more closely with the networking
stack's understanding of the traffic transiting the kernel.

This safety is provided by tracking pointer values through each conditional
path in the program, and rejecting the load of programs that fail to "balance"
resource acquisition (lookup) with release.

\section{Background}

The BPF verifier must ensure that all BPF programs that are loaded into the
kernel are safe to execute. When accessing kernel resources, this includes
providing guarantees around the availability of the underlying resource for the
duration of the BPF program. For some socket types, such as UDP sockets, or TCP
listen sockets, this can be achieved with minimal implementation: The
destruction of such sockets is already governed by standard RCU rules, meaning
that while the RCU lock is held, they can be safely referenced; Once the RCU
grace period is reached, the memory may be freed and references to the socket
are no longer safe to use. Given that BPF programs run under the RCU lock, this
allows simple, safe access of properties of these sockets without requiring
additional mechanisms to ensure the safety of the socket pointers.

When it comes to TCP sockets that are not governed by RCU, the standard
refcounting model is used instead. To ensure that the socket remains safe to
access for a period of BPF program execution, the BPF program must also take a
reference on the socket. Furthermore, if references must be taken, then they
also must be released.

@@

@@

@@

Alternatives:
\begin{itemize}
\item RCU-protect other socket types
\item Collect references during program execution and release afterwards
\item Build acquire/release semantics into the API
\end{itemize}

Collecting references and releasing afterwards has the unfortunate overhead
that even the execution of BPF programs that do not make use of the new socket
lookup helpers would need to pay the cost of checking the list of references
to free. When dealing with sufficiently demanding use cases such as those which
use XDP, even the few instructions required to implement such a check may have
a measurable impact on performance.

Calls to helper functions which acquire references to resources are annotated
in the verifier to associate the acquisition and release functions. When the
instruction that acquires a resource is processed, the verifier state is
updated to include a resource identifier for the resulting value, and the
register's value is associated with this identifier. When processing a
corresponding release helper call, if a parameter to the function contains a
resource identifier, then the corresponding identifire is removed from the
verifier state. If any paths in the program reach the final (\verb+BPF_EXIT+)
instruction and the verifier state still contains resource identifiers, then
the program is considered unsafe as it has leaked the resource.

@@

\todo{LD\_ABS, LD\_IND}

@@

@@

This paper describes the contribution of two extensions to the Linux kernel BPF
implementation, firstly the introduction of reference tracking logic into the
verifier, which provides useful base infrastructure for supporting acquire and
release semantics for kernel resources, and secondly the socket lookup API
which may be now used from packet paths to allow BPF writers to build logic
based upon introspection of kernel sockets.

@@

The core BPF maintainers for their guidance and review of the patches.
